# Identification problem and standard identification techniques {#identifStruct}

## The identification problem {#IdentifPbm}

In Section \@ref(estimVAR), we have seen how to estimate $\mathbb{V}ar(\varepsilon_t) =\Omega$ and the $\Phi_k$ matrices in the context of a VAR model. But the IRFs are functions of $B$ and of the $\Phi_k$'s, not of $\Omega$ the $\Phi_k$'s (see Section \@ref(IRFSVARMA)). We have $\Omega = BB'$, which provides some restrictions on the components of $B$, but this is not sufficient to fully identify $B$. Indeed, seen a system of equations whose unknowns are the $b_{i,j}$'s (components of $B$), the system  $\Omega = BB'$ contains only $n(n+1)/2$ linearly independent equations. For instance, for $n=2$:
\begin{eqnarray*}
&&\left[
\begin{array}{cc}
\omega_{11} & \omega_{12} \\
\omega_{12} & \omega_{22}
\end{array}
\right] = \left[
\begin{array}{cc}
b_{11} & b_{12} \\
b_{21} & b_{22}
\end{array}
\right]\left[
\begin{array}{cc}
b_{11} & b_{21} \\
b_{12} & b_{22}
\end{array}
\right]\\
&\Leftrightarrow&\left[
\begin{array}{cc}
\omega_{11} & \omega_{12} \\
\omega_{12} & \omega_{22}
\end{array}
\right] = \left[
\begin{array}{cc}
b_{11}^2+b_{12}^2 & \color{red}{b_{11}b_{21}+b_{12}b_{22}} \\
\color{red}{b_{11}b_{21}+b_{12}b_{22}} & b_{22}^2 + b_{21}^2
\end{array}
\right].
\end{eqnarray*}

We then have 3 linearly independent equations but 4 unknowns. Therefore, $B$ is not identified based on second-order moments. Additional restrictions are required to identify $B$. This section covers two standard identification schemes: **short-run** and **long-run** restrictions.

1. A **short-run restriction (SRR)** prevents a structural shock from affecting an endogenous variable contemporaneously.
* It is easy to implement: the appropriate entries of $B$ are set to 0.
* A particular (popular) case is that of the **Cholesky, or recursive approach**.
* Examples include @BERNANKE198649, @Sims_1986, @Gali_1992, @RubioRamirez_et_al_2010.

2. A **long-run restriction (LRR)** prevents a structural shock from having a cumulative impact on one of the endogenous variables. 
* Additional computations are required to implement this. One needs to compute the cumulative effect of one of the structural shocks $u_{t}$ on one of the endogenous variable.
* Examples include @Blanchard_Quah_1989, @Faust_Leeper_1997, @Gali_1999, @Erceg_et_al_2005, @NBERc11177.

The two approaches can be combined (see, e.g., @Gerlach_Smets_1995).


## A stylized example motivating short-run restrictions

Let us consider a simple example that could motivate short-run restrictions. Consider the following stylized macro model:
\begin{equation}
\begin{array}{clll}
g_{t}&=& \bar{g}-\lambda(i_{t-1}-\mathbb{E}_{t-1}\pi_{t})+ \underbrace{{\color{blue}\sigma_d \eta_{d,t}}}_{\mbox{demand shock}}& (\mbox{IS curve})\\
\Delta \pi_{t} & = & \beta (g_{t} - \bar{g})+ \underbrace{{\color{blue}\sigma_{\pi} \eta_{\pi,t}}}_{\mbox{cost push shock}} & (\mbox{Phillips curve})\\
i_{t} & = & \rho i_{t-1} + \left[ \gamma_\pi \mathbb{E}_{t}\pi_{t+1}  + \gamma_g (g_{t} - \bar{g}) \right]\\
&& \qquad \qquad+\underbrace{{\color{blue}\sigma_{mp} \eta_{mp,t}}}_{\mbox{Mon. Pol. shock}} & (\mbox{Taylor rule}),
\end{array}(\#eq:systemI)
\end{equation}
where:
\begin{equation}
\eta_t = 
\left[
\begin{array}{c}
\eta_{\pi,t}\\
\eta_{d,t}\\
\eta_{mp,t}
\end{array}
\right]
\sim i.i.d.\,\mathcal{N}(0,I).(\#eq:covU)
\end{equation}

Vector $\eta_t$ is assumed to be a vector of structural shocks, mutually and serially independent. On date $t$:

* $g_t$ is contemporaneously affected by $\eta_{d,t}$ only;
* $\pi_t$ is contemporaneously affected by $\eta_{\pi,t}$ and $\eta_{d,t}$;
* $i_t$ is contemporaneously affected by $\eta_{mp,t}$, $\eta_{\pi,t}$ and $\eta_{d,t}$.

System \@ref(eq:systemI) could be rewritten as follows:
\begin{equation}
\left[\begin{array}{c}
g_t\\
\pi_t\\
i_t
\end{array}\right]
= \Phi(L)
\left[\begin{array}{c}
g_{t-1}\\
\pi_{t-1}\\
i_{t-1} +
\end{array}\right] +\underbrace{\underbrace{
\left[
\begin{array}{ccc}
0 & \bullet & 0 \\
\bullet & \bullet & 0 \\
\bullet & \bullet & \bullet
\end{array}
\right]}_{=B} \eta_t.}_{=\varepsilon_t}(\#eq:BBBB)
\end{equation}

This is the **reduced-form** of the model. This representation suggests three additional restrictions on the entries of $B$; the latter matrix is therefore identified as soon as $\Omega = BB'$ is known (up to the signs of its columns).


## Cholesky: a specific short-run-restriction situation {#Cholesky}

There are particular cases in which some well-known matrix decomposition of $\Omega=\mathbb{V}ar(\varepsilon_t)$ can be used to easily estimate some specific SVAR. This is the case for the so-called Cholesky decomposition. Consider the following context:

* A first shock (say, $\eta_{n_1,t}$) can affect instantaneously
(i.e., on date $t$) only one of the endogenous variable (say, $y_{n_1,t}$);
* A second shock (say, $\eta_{n_2,t}$) can affect instantaneously
(i.e., on date $t$) two endogenous variables, $y_{n_1,t}$ (the same as before) and  $y_{n_2,t}$;
* $\dots$

This implies

1. that column $n_1$ of $B$ has only 1 non-zero entry (this is the $n_1^{th}$ entry),
2. that column $n_2$ of $B$ has 2 non-zero entries (the $n_1^{th}$ and the $n_2^{th}$ ones), etc.

Without loss of generality, we can set $n_1=n$, $n_2=n-1$, etc. In this context, matrix $B$ is lower triangular. The Cholesky decomposition of $\Omega_{\varepsilon}$ then provides an appropriate estimate of $B$, since this matrix decomposition yields to a lower triangular matrix satisfying:
$$
\Omega_\varepsilon = BB'.
$$

For instance, @DEDOLA20051543 estimate 5 structural VAR models for the US, the UK, Germany, France and Italy to analyse the monetary-policy transmission mechanisms. They estimate SVAR(5) models over the period 1975-1997. The shock-identification scheme is based on Cholesky decompositions, the ordering of the endogenous variables being: (1) the industrial production, (2) the consumer price index, (3) a commodity price index, (4) the short-term rate, (5) monetary aggregate and (6) the effective exchange rate (except for the US). This ordering implies that monetary policy (i.e., the short-term rate) reacts to the shocks affecting the first three variables but that the latter react to monetary policy shocks with a one-period lag only.

Consider again the small structural example discussed above, whose reduced-form VAR representation is given in Eq. \@ref(eq:BBBB). If we reorder the structural shocks as $\eta_t = (\eta_{d,t},\eta_{\pi,t},\eta_{mp,t})'$ (the order being arbitrary), the impact matrix $B$ can be written as
$$
B = \begin{bmatrix}
\bullet & 0 & 0 \\
\bullet & \bullet & 0 \\
\bullet & \bullet & \bullet
\end{bmatrix},
$$
that is, $B$ corresponds to the Cholesky factor of $\Omega$.

However, it is not always possible to obtain a lower-triangular $B$ matrix by reordering the structural shocks. For instance, if for a given ordering we have
$$
B = \begin{bmatrix}
0 & \bullet & \bullet \\
\bullet & 0 & \bullet \\
\bullet & \bullet & 0
\end{bmatrix},
$$
then no reordering of the shocks will produce a lower-triangular structure.



The Cholesky approach can be employed when one is interested in one specific structural shock. This is the case, e.g., of @Christiano_Eichenbaum_Evans_1996. Their identification is based on the following relationship between $\varepsilon_t$ and $\eta_t$:
$$
\left[\begin{array}{c}
\boldsymbol\varepsilon_{S,t}\\
\varepsilon_{r,t}\\
\boldsymbol\varepsilon_{F,t}
\end{array}\right] =
\left[\begin{array}{ccc}
B_{SS} & 0 & 0 \\
B_{rS} & B_{rr} & 0 \\
B_{FS} & B_{Fr} & B_{FF}
\end{array}\right]
\left[\begin{array}{c}
\boldsymbol\eta_{S,t}\\
\eta_{r,t}\\
\boldsymbol\eta_{F,t}
\end{array}\right],
$$
where $S$, $r$ and $F$ respectively correspond to *slow-moving variables*, the policy variable (short-term rate) and *fast-moving variables*. While $\eta_{r,t}$ is scalar, $\boldsymbol\eta_{S,t}$ and $\boldsymbol\eta_{F,t}$ may be vectors. The space spanned by $\boldsymbol\varepsilon_{S,t}$ is the same as that spanned by $\boldsymbol\eta_{S,t}$. As a result, because $\varepsilon_{r,t}$ is a linear combination of $\eta_{r,t}$ and $\boldsymbol\eta_{S,t}$ (which are $\perp$), it comes that the $B_{rr}\eta_{r,t}$'s are the (population) residuals in the regression of $\varepsilon_{r,t}$ on $\boldsymbol\varepsilon_{S,t}$. Because $\mathbb{V}ar(\eta_{r,t})=1$, $B_{rr}$ is given by the square root of the variance of  $B_{rr}\eta_{r,t}$. $B_{F,r}$ is finally obtained by regressing the components of $\boldsymbol\varepsilon_{F,t}$ on the estimates of $\eta_{r,t}$.

An equivalent approach consists in computing the Cholesky decomposition of $BB'$ and the contemporaneous impacts of the monetary policy shock (on the $n$ endogenous variables) are the components of the column of $B$ corresponding to the policy variable.


```{r CEE, fig.align = 'left-aligned', out.width = "95%", fig.cap = "Response to a monetary-policy shock. Identification approach of Christiano, Eichenbaum and Evans (1996). Confidence intervals are obtained by boostrapping the estimated VAR model (see inference section).",warning=FALSE,message=FALSE}
library(IdSS)
library(vars)
data("USmonthly")
# Select sample period:
First.date <- "1965-01-01";Last.date <- "1995-06-01"
indic.first <- which(USmonthly$DATES==First.date)
indic.last  <- which(USmonthly$DATES==Last.date)
USmonthly   <- USmonthly[indic.first:indic.last,]
considered.variables <- c("LIP","UNEMP","LCPI","LPCOM","FFR","NBR","TTR","M1")
y <- as.matrix(USmonthly[considered.variables])
res.svar.ordering <- svar.ordering(y,p=3,
                                   posit.of.shock = 5,
                                   nb.periods.IRF = 20,
                                   nb.bootstrap.replications = 100,
                                   confidence.interval = 0.90, # expressed in pp.
                                   indic.plot = 1 # Plots are displayed if = 1.
                                   )
```


## Long-run restrictions

Let us now turn to **long-run restrictions**. Such a restriction concerns the long-run influence of a shock on an endogenous variable. Let us consider for instance a structural shock that is assumed to have no "long-run influence" on GDP. How to express this? The long-run change in GDP can be expressed as $GDP_{t+h} - GDP_t$, with $h$ large. Note further that:
$$
GDP_{t+h} - GDP_t = \Delta GDP_{t+h} +\Delta GDP_{t+h-1} + \dots + \Delta GDP_{t+1}.
$$
Hence, the fact that a given structural shock ($\eta_{i,t}$, say) has no long-run influence on GDP means that
$$
\lim_{h\rightarrow\infty}\frac{\partial GDP_{t+h}}{\partial \eta_{i,t}} = \lim_{h\rightarrow\infty} \frac{\partial}{\partial \eta_{i,t}}\left(\sum_{k=1}^h \Delta  GDP_{t+k}\right)= 0.
$$

This long-run effect can be formulated as a function of $B$ and of the matrices $\Phi_i$ when $y_t$ (including $\Delta GDP_t$) follows a VAR process.

### The VAR(1) case

Let us start with the VAR(1) case. We have:
\begin{eqnarray*}
y_{t} &=& c+\Phi y_{t-1}+\varepsilon_{t}\\
& = & c+\varepsilon_{t}+\Phi(c+\varepsilon_{t-1})+\ldots+\Phi^{k}(c+\varepsilon_{t-k})+\ldots \\
& = & \mu +\varepsilon_{t}+\Phi\varepsilon_{t-1}+\ldots+\Phi^{k}\varepsilon_{t-k}+\ldots \\
& = & \mu +B\eta_{t}+\Phi B\eta_{t-1}+\ldots+\Phi^{k}B\eta_{t-k}+\ldots,
\end{eqnarray*}
which is the Wold representation of $y_t$.

The sequence of shocks $\{\eta_t\}$ determines the sequence $\{y_t\}$. What if $\{\eta_t\}$ is replaced with $\{\tilde{\eta}_t\}$, where $\tilde{\eta}_t=\eta_t$ if $t \ne s$ and $\tilde{\eta}_s=\eta_s + \gamma$? Assume $\{\tilde{y}_t\}$ is the associated "perturbated" sequence. We have $\tilde{y}_t = y_t$ if $t<s$. For $t \ge s$, the Wold decomposition of $\{\tilde{y}_t\}$ implies:
$$
\tilde{y}_t = y_t + \Phi^{t-s} B \gamma.
$$
Therefore, the cumulative impact of $\gamma$ on $\tilde{y}_t$ will be (for $t \ge s$):
\begin{eqnarray}
(\tilde{y}_t - y_t) +  (\tilde{y}_{t-1} - y_{t-1}) + \dots +  (\tilde{y}_s - y_s) &=& \nonumber \\
(Id + \Phi + \Phi^2 + \dots + \Phi^{t-s}) B \gamma.&& (\#eq:cumul)
\end{eqnarray}

Consider a shock on $\eta_{1,t}$, with a magnitude of $1$. This shock corresponds to $\gamma = [1,0,\dots,0]'$. Given Eq. \@ref(eq:cumul), the long-run cumulative effect of this shock on the endogenous variables is given by:
$$
\underbrace{\underbrace{(Id+\Phi+\ldots+\Phi^{k}+\ldots)}_{=(Id - \Phi)^{-1}}B}_{=: \Theta}\left[\begin{array}{c}
1\\
0\\
\vdots\\
0\end{array}\right],
$$
which is the first column of the $n \times n$ matrix $\Theta := (Id - \Phi)^{-1}B$.

In this context, consider the following long-run restriction: *"the $j^{th}$ structural shock has no cumulative impact on the $i^{th}$ endogenous variable"*. It is equivalent to
$$
\Theta_{ij}=0,
$$
where $\Theta_{ij}$ is the element $(i,j)$ of $\Theta$.

If $n(n-1)/2$ restrictions of this type can be made, then $B$ is identified. In particular, in the case case where $\Theta$ is lower-triangular, the problem admits an analytical solution. Indeed, let $J = (I - \Phi)^{-1}$ (assumed to be invertible). We want $\Theta = JB$ to be lower triangular and $\Omega = BB'$. Since $B = J^{-1} \Theta$, we have
\[
\Omega = BB' = J^{-1} \Theta \Theta' (J^{-1})' \;\Rightarrow\; J\Omega J' = \Theta \Theta'.
\]
Since $\Omega$ is positive definite and $J$ is invertible, $\Sigma = J\Omega J'$ is positive definite. Take the (unique, with positive diagonal) Cholesky factorization $\Sigma = \Theta \Theta'$ with $\Theta$ lower triangular. Then set $B = J^{-1} \Theta$. This $B$ satisfies $\Omega = BB'$ and $(I - \Phi)^{-1}B = \Theta$ is lower triangular.


### The VAR($p$) case

Several of the developments made above are still valid in the VAR($p$) case since a VAR($p$) process can be rewritten as a VAR(1) process by augmenting the state vector. More specifically, stack the last $p$ values of vector $y_t$ in vector $y_{t}^{*}=[y_t',\dots,y_{t-p+1}']'$; Eq. \@ref(eq:yVAR) can then be rewritten in its **companion form**:
\begin{equation}
y_{t}^{*} =
\underbrace{\left[\begin{array}{c}
c\\
0_{n \times 1}\\
\vdots\\
0_{n \times 1}\end{array}\right]}_{=c^*}+
\underbrace{\left[\begin{array}{cccc}
\Phi_{1} & \Phi_{2} & \cdots & \Phi_{p}\\
I & 0 & \cdots & 0\\
0 & \ddots & 0 & 0\\
0 & 0 & I & 0\end{array}\right]}_{=\Phi}
y_{t-1}^{*}+
\underbrace{\left[\begin{array}{c}
\varepsilon_{t}\\
0_{n \times 1}\\
\vdots\\
0_{n \times 1}\end{array}\right]}_{\varepsilon_t^*},(\#eq:ystarVAR)
\end{equation}
where matrices $\Phi$ and $\Omega^*:= \mathbb{V}ar(\varepsilon_t^*)$ are of dimension $np \times np$. Matrix $\Phi$ had been introduced in Eq. \@ref(eq:matrixPHI), and
$$
\Omega^* := \mathbb{V}ar(\varepsilon_t^*)=
\left[\begin{array}{c}
B \\
0_{n \times n}  \\
\vdots \\
0_{n \times n}
\end{array}\right]\left[\begin{array}{cccc}
B & 0_{n \times n} & \dots & 0_{n \times n}
\end{array}\right]
= \left[\begin{array}{cccc}
\Omega & 0_{n \times n} & \dots & 0_{n \times n} \\
0_{n \times n} & 0_{n \times n} \\
\vdots &&\ddots\\
0_{n \times n} & \dots &&0_{n \times n}
\end{array}\right].
$$
In that context, the long-run effect on $y_t^*+y_{t+1}^*+y_{t+2}^*+\dots$ of a change in $\eta_t$ by $\gamma$ (that happens on date $t$) is given by:
$$
\underbrace{(Id+\Phi+\ldots+\Phi^{k}+\ldots)}_{=(Id - \Phi)^{-1}}\left[\begin{array}{c}
B \\
0_{n \times n}  \\
\vdots \\
0_{n \times n}
\end{array}\right]\gamma.
$$
In particular, the cumulated effect of this shock on $y_t$ (and not $y_t^*$ anymore) is $J B \gamma$, where $J$ is the upper-left $n \times n$ submatrix of $(I - \Phi)^{-1}$. Let us denote the matrix $J B$ by $\Theta$. With these notations, the long-run restriction: *the $j^{th}$ structural shock has no cumulative impact on the $i^{th}$ endogenous variable* is equivalent to the fact that the component $(i,j)$ of $\Theta$ is equal to zero.

If $n(n-1)/2$ restrictions of this type can be made, then $B$ is identified. In particular, in the case where $\Theta$ is lower triangular, the problem admits an analytical solution. Using $B = J^{-1} \Theta$, we get
\[
\Omega = B B' = J^{-1} \Theta \Theta' (J^{-1})' \;\Rightarrow\; J \Omega J' = \Theta \Theta'.
\]
Since $\Omega$ is positive definite, $\Sigma = J \Omega J'$ is positive definite. Take the (unique, with positive diagonal) Cholesky factorization $\Sigma = \Theta \Theta'$ with $\Theta$ lower triangular. Then set $B = J^{-1} \Theta$. This $B$ satisfies $\Omega = B B'$ and $J B = \Theta$ is lower triangular.


### Blanchard and Quah (1989)

@Blanchard_Quah_1989 have implemented such long-run restrictions in a small-scale VAR. Two variables are considered: GDP and unemployment. Consequently, the VAR is affected by two types of shocks. Specifically, authors want to identify **supply shocks** (that can have a permanent effect on output) and **demand shocks** (that cannot have a permanent effect on output).[^Footnote:BQstruct]

[^Footnote:BQstruct]: The motivation of the authors regarding their long-run restrictions can be obtained from a traditional Keynesian view of fluctuations. The authors propose a variant of a model from @Fischer_1977.
<!-- \begin{eqnarray} -->
<!-- Y_{t} & = & M_{t}-P_{t}+a.\theta_{t}(\#eq:demand)\\ -->
<!-- Y_{t} & = & N_{t}+\theta_{t}(\#eq:prodfunct)\\ -->
<!-- P_{t} & = & W_{t}-\theta_{t}(\#eq:PS)\\ -->
<!-- W_{t} & = & W\mid\left\{ \mathbb{E}_{t-1}N_{t}=\overline{N}\right\}. (\#eq:WS) -->
<!-- \end{eqnarray} -->
<!-- To close the model, the authors assume the following dynamics for the money supply and the productivity: -->
<!-- \begin{eqnarray*} -->
<!-- M_{t} & = & M_{t-1}+\varepsilon_{t}^{d}\\ -->
<!-- \theta_{t} & = & \theta_{t-1}+\varepsilon_{t}^{s}. -->
<!-- \end{eqnarray*} -->
<!-- In this context, it can be shown that -->
<!-- \begin{eqnarray*} -->
<!-- \Delta Y_{t} & = & (\varepsilon_{t}^{d}-\varepsilon_{t-1}^{d})+a.(\varepsilon_{t}^{s}-\varepsilon_{t-1}^{s})+\varepsilon_{t}^{s}\\ -->
<!-- u_{t} & = & -\varepsilon_{t}^{d}-a\varepsilon_{t}^{s} -->
<!-- \end{eqnarray*} -->
<!-- Then, it appears that the demand shocks have no long-run cumulative impact on $\Delta Y_{t}$, the GDP growth, i.e. no long-term impact on output $Y_t$. The vector of endogenous variables is $y_t = [\Delta Y_{t} \quad u_{t}]'$ where $\Delta Y_{t}$ denotes the GDP growth. -->

@Blanchard_Quah_1989's dataset is quarterly, spanning the period from 1950:2 to 1987:4. Their VAR features 8 lags. Here are the data they use:

```{r BQ1, fig.align = 'left-aligned', out.width = "95%", fig.cap = "These data come from Blanchard and Quah (1989). GDP growth rates are calculated as the first differences of the logarithm of GDP."}
library(IdSS)
data(BQ)
par(mfrow=c(1,2))
plot(BQ$Date,BQ$Dgdp,type="l",main="GDP quarterly growth rate",
     xlab="",ylab="",lwd=2)
plot(BQ$Date,BQ$unemp,type="l",ylim=c(-3,6),main="Unemployment rate (gap)",
     xlab="",ylab="",lwd=2)
```

Estimate a reduced-form VAR(8) model:

```{r BQ2, warning=FALSE, message=FALSE}
library(vars)
y <- BQ[,2:3]
n <- dim(y)[2]
p <- 8
est.VAR <- VAR(y,p=p)
Omega <- var(residuals(est.VAR))
```

Let us employ the approach described above:

```{r}
Phi <- Acoef(est.VAR)
PHI <- make.PHI(Phi)
A <- diag(n*p) - PHI
J <- solve(A)[1:n,1:n]
Sigma <- J %*% Omega %*% t(J)
L <- t(chol(Sigma))
B <- solve(J) %*% L
```

Now, let us define a loss function (`loss`) that is equal to zero if (a) $BB'=\Omega$ and (b) the element (1,1) of $\Theta = (Id - \Phi)^{-1} B$ is equal to zero:

```{r BQ3}
# Compute (Id - Phi)^{-1}:
Phi <- Acoef(est.VAR)
PHI <- make.PHI(Phi)
sum.PHI.k <- solve(diag(dim(PHI)[1]) - PHI)[1:2,1:2]
loss <- function(param){
  B <- matrix(param,2,2)
  X <- Omega - B %*% t(B)
  Theta <- sum.PHI.k[1:2,1:2] %*% B
  loss <- 10000 * ( X[1,1]^2 + X[2,1]^2 + X[2,2]^2 + Theta[1,1]^2 )
  return(loss)
}
res.opt <- optim(c(1,0,0,1),loss,method="BFGS",hessian=FALSE)
print(res.opt$par)
```

(Note: one can use that type of approach, based on a loss function, to mix short- and long-run restrictions.) 

Figure \@ref(fig:BQ4) displays the resulting IRFs. Note that, for GDP, we cumulate the GDP growth IRF, so as to have the response of the GDP in level.

```{r BQ4, fig.align = 'left-aligned', out.width = "95%", fig.cap = "IRF of GDP and unemployment to demand and supply shocks."}
B.hat <- matrix(res.opt$par,2,2)
print(cbind(Omega,B.hat %*% t(B.hat)))
nb.sim <- 40
par(mfrow=c(2,2));par(plt=c(.15,.95,.15,.8))
Y <- simul.VAR(c=matrix(0,2,1),Phi,B.hat,nb.sim,y0.star=rep(0,2*8),
               indic.IRF = 1,u.shock = c(1,0))
plot(cumsum(Y[,1]),type="l",lwd=2,xlab="",ylab="",main="Demand shock on GDP")
plot(Y[,2],type="l",lwd=2,xlab="",ylab="",main="Demand shock on UNEMP")
Y <- simul.VAR(c=matrix(0,2,1),Phi,B.hat,nb.sim,y0.star=rep(0,2*8),
               indic.IRF = 1,u.shock = c(0,1))
plot(cumsum(Y[,1]),type="l",lwd=2,xlab="",ylab="",main="Supply shock on GDP")
plot(Y[,2],type="l",lwd=2,xlab="",ylab="",main="Supply shock on UNEMP")
```

